Long-term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-07-04 article/article.Rmd
Commit 2024-07-05 00:14

First, we evaluated the performance of the modelled $\text{GHI}_\text{ref}$ in
relation to the measured GHI for different levels of atmospheric clearness, by using
in the RTM the monthly climatological AOD and WC, less their respective standard
deviations.
<!--
$\tau_{\text{cs}} = \tau_{500\text{nm}} - 1\sigma$ and $w_{h\text{cs}} = w_h - 1\sigma$
($\text{GHI}_\text{CSm}$).
-->
These values represent typical atmospheres in Thessaloniki with lower than average
load of aerosols and humidity, which are the main factors that attenuate the GHI,
excluding clouds.  With this approach the simulated $\text{GHI}_\text{ref}$ should be
generally greater than the measured GHI when aerosols are more abundant.  To
compensate for this, we defined the following threshold $E$ to compare the
measured $\text{GHI}$: 
<!--
($`r 100*(C4_cs_ref_ratio-1)`\,\%$)
with an additional constant offset of
$`r C4_GLB_diff_THRES`\,\text{W}/\text{m}^2$,
as described in Equation\nobreakspace\ref{eq:CE4}.
-->
\begin{equation}
\text{CE} : E > `r C4_GLB_diff_THRES` + `r C4_cs_ref_ratio` \cdot \text{GHI}_\text{ref} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation}
This is the criterion of our CE identification.

@@

These values were determined through the implementation of an empirical method with
manual inspection of the CE identification results on selected days of the whole
dataset.  We used seven sets of days with characteristics relevant to the efficiency
of the identification threshold.  These sets were random groups of about 20 -- 30
days with the following characteristics:
(a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance,
(c) absence of clouds (by implementing a clear sky identification algorithm as discussed in @Natsis2023),
(d) absence of clouds and absence of EC events,
(e) with at least $60\,\%$ of the day length without clouds and presence of EC events,
(h) randomly selected days, and
(i) manually selected days.
For the latter case and where needed, we also used images from a sky-camera to
further aid the decision of the manual inspection.

@@

The definition of the CE events with this method has a degree of subjectivity, since
the actual clear sky irradiance is not known and can only be approximated.  However,
this method is capable in detecting all major CE events.  Where some CE events with
very low OIR may be not detected, these are few with small over-irradiance, and it is
unlikely that will affect significantly our results.

@@

```{r example-day, echo=FALSE, fig.cap="Example of CE identification for 2019-07-11. The green line with blue symbols depicts the measured GHI in one minute steps. The red line shows the modelled threshold $E$ for the detection of CE events, which are denoted with red circles. The black curve represents the TSI at the top of the atmosphere, adjusted for the actual Sun-Earth distance and multiplied by the cosine of the cosine of SZA.", fig.pos='H'}
knitr::include_graphics("../images/example-days-18.png")
```

```{r example-year, echo=FALSE, fig.cap="Example scatter plot of the measured GHI and the reference clear sky irradiance for the year 2005. The over-irradiance for CE and ECE events is color coded, while the remaining data points are shown in black.", fig.pos='H'}
knitr::include_graphics("../images/P-example-years-13.png")
```

@@

<!-- \FloatBarrier -->

@@

The main focus of this study is to investigate the time evolution of the CE events by
analyzing the GHI measurements at Thessaloniki. Cloud enhancements can be influenced
by different factors, such as the geometry, size and optical thickness of clouds,
their height in the atmosphere and on local weather regimes [@Mol2023; @Veerman2022;
@Gristey2022; @Tzoumanikas2016].  Some of these factors are related to changes in
climate; hence it would be reasonable to expect capturing their contributions to the
frequency of occurrence of CE events over Thessaloniki, as well as to the average OIR
and excess irradiation.  The long-term trends were calculated using a first-order
autoregressive model with the 'maximum likelihood' fitting method [@Gardner1980;
@Jones1980], by implementing the function 'arima' from the library 'stats' of the R
programming language [@RCT2023]. All trends are reported together with their
$2\sigma$ error.

@@

We have to note that the excess irradiation related to the CE events can not be
directly linked to the total energy balance of the atmosphere. The net solar
radiation of the region is not increased, but is rather redistributed through the CE
events.  This is also depicted by the ECE irradiance values, which can exceed the
equivalent clear sky irradiance by a significant amount.

@@

Another interesting aspect of the CE events is their distribution within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the monthly box and
whisker plot of the CE number of occurrence normalized with the highest median value,
that occurs in June, depicting a clear seasonal cycle.  Although CE events are
present throughout the year, the most active months are May and June. During the
winter (December -- February), the number of CE cases is about $25\,\%$ of the
maximum, while in the intermediate months, the number of occurrences gradually ramps
between the maximum and minimum.  This seasonality is a combined effect of different
factors, among them the types of clouds, their frequency of occurrence, the
seasonally varying relative position of the sun, as well as the local landscape
characteristics that may influence the formation of the clouds.  Unfortunately, lack
of detailed data on cloud formation, type and location is not allowing further
analysis. The interannual variability of the monthly CE events is quite high as
manifested by the large monthly extremes, especially in the summer.

```{r relative-month-occurrences, echo=FALSE, fig.cap="Seasonal variability of the number of CE events normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have been also normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\\,\\%$ and the upper $75\\,\\%$ percentiles, the thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values and the solid circles are outliers."}
knitr::include_graphics("../images/clim-CE-month-norm-MAX-median-N-MW-1.png")
```

@@

```{r ovir-distribution, echo=FALSE, fig.cap="Relative frequency distribution of CE events as a function of OIR."}
knitr::include_graphics("../images/P-relative-distribution-diff-1.png")
```

@@

The duration of the CE events is variable and can last for several minutes or even
more than an hour. In order to study the characteristics of these consecutive events,
we grouped them into bins of increasing duration in steps of one minute.  We have
identified `r ST_G0[, .N]` groups of CE in the whole period of study, where the group
of the longest duration of
`r ST_G0[, max(GLB_diff.N)]` minutes occurred on
`r strftime(ST_G0[which.max(GLB_diff.N), Date], "%d\\ %B %Y")`.
Figure\nobreakspace{}\ref{fig:ceg-duration-distribution} shows the frequency
distribution of the CE events according to their duration.  We conclude that although
some groups of events last for more than an hour about $80\,\%$ have duration of less
than 5 minutes.

```{r ceg-duration-distribution, echo=FALSE, fig.cap="Relative frequency distribution of CE groups of consequent CE events according to their duration."}
knitr::include_graphics("../images/groups-1.png")
```

The relation between the duration and the mean OIR of the groups has also been
studied (Figure\nobreakspace{}\ref{fig:group-2d}).  Evidently, evens of high excess
irradiation have small duration and vice versa. The vast majority of grouped events
are associated with small excess irradiation (e.g. $<5\,\text{kJ}/\text{m}^2$) and
small duration (e.g. $<5\,\text{min}$) while groups with strong excess irradiation
and long duration are very rare.  Similar results of this relation have been
reported by @Zhang2018, on a study using a far higher sampling rate than ours.

```{r group-2d, echo=FALSE, fig.cap="Relation of excess irradiation of CE groups with their duration. The logarithmic color scale denotes the frequency of the respective groups of events."}
knitr::include_graphics("../images/P-groups-bin2d-1.png")
```

@@

An aspect of the CE events that is commonly reported and has some significance
on the solar energy production infrastructure are the extreme CE events (ECE). Where
solar irradiance exceeds the expected irradiance on top of the atmosphere
(Equation\nobreakspace{}\ref{eq:ECE}).
Analogous to Figure\nobreakspace{}\ref{fig:relative-month-occurrences} we have
computed the distribution of the number of occurrences of ECE events by month in
Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}. The most active period
for ECE events is in the spring and the start of the summer (March -- June), followed
by a period in the late fall (September and October).  This is probably related to 
the weather characteristics of these periods, where there are continuous alternations
between clear sky periods and clouds.

```{r relative-month-occurancies-ECE, echo=FALSE, fig.cap="Seasonal variability of the number of ECE events normalized to the maximum occurring in March, in the form of a box and whisker plot. The box contains the data between the lower $25\\,\\%$ and the upper $75\\,\\%$ percentiles. The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values."}
knitr::include_graphics("../images/clim-ECE-month-norm-MAX-median-N-2.png")
```

@@

The distribution of the ECE events (Figure\nobreakspace{}\ref{fig:P-extreme-distribution}),
shows that there are rare cases where the OIR can exceed the TSI even more than
$400\,\text{W}/\text{m}^2$, with the $75\,\%$ of the cases to be below
$200\,\text{W}/\text{m}^2$. 
Those finds are in accordance with results form @Vamvakas2020, 
with the difference that the OIR values are lower for Thessaloniki.

```{r P-extreme-distribution, echo=FALSE, fig.cap="Distribution of ECE above clear sky threshold, for cases that are exceeding the TSI."}
knitr::include_graphics("../images/P-extreme-distribution-1.png")
```

@@

By creating a clear sky approximation of the GHI, which represents the long- and
short-term variation of the expected clear sky GHI, we were able to identify cases of 
CE events.
After analyzing the CE cases, we found an increase of 
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the
mean annual total energy of the CE events increasing with a rate of
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The most active season of CE events over Thessaloniki is concentrated on early
summer, on May and June.
We found that continuous CE events can last up to $140$ minutes, and $80\,\%$ of
them bellow $5$ minutes.

The magnitude of the ECE events does not exceed the values reported for sites with
more favourable conditions for the phenomenon. We have observed cases exceeding the
TSI by $400\,\text{W}/\text{m}^{2}$ with $75\,\%$ of the cases under
$200\,\text{W}/\text{m}^{2}$. The climatological characteristic of the ECE events,
showed that the most active months are spread in half of the year (March -- June,
September and October). We found that CE conditions, can have a duration of more than
an hour in rare cases, with the bulk of the cases having a duration under 5 minutes.
Some of the characteristics of CE and ECE events we analysed have strong similarities
with results by @Vamvakas2020 for the city of Patras, south of Thessaloniki, with
analog differences on the intensity of the solar radiation.

An interpretation of the CE trends shows that the interaction of GHI with the clouds,
through this 30-year period, is a dynamic phenomenon that needs further
investigation.  Although, to approach it, we need more long-term observations of AOD
and clouds characteristics.

>>>>

First, we evaluated the performance of the modelled $\text{GHI}_\text{ref}$ in
relation to the measured GHI for different levels of atmospheric clearness, by using
in the RTM the monthly climatological AOD and WC, less their respective standard
deviations.
<!--
$\tau_{\text{cs}} = \tau_{500\text{nm}} - 1\sigma$ and $w_{h\text{cs}} = w_h - 1\sigma$
($\text{GHI}_\text{CSm}$).
-->
These values represent typical atmospheres in Thessaloniki with lower than average
load of aerosols and humidity, which are the main factors that attenuate the GHI,
excluding clouds.  With this approach the simulated $\text{GHI}_\text{ref}$ should be
generally greater than the measured GHI when aerosols are more abundant.  To
compensate for this, we defined the following threshold $E$ to compare the
measured $\text{GHI}$: 
<!--
($`r 100*(C4_cs_ref_ratio-1)`\,\%$)
with an additional constant offset of
$`r C4_GLB_diff_THRES`\,\text{W}/\text{m}^2$,
as described in Equation\nobreakspace\ref{eq:CE4}.
-->
\begin{equation}
\text{CE} : E > `r C4_GLB_diff_THRES` + `r C4_cs_ref_ratio` \cdot \text{GHI}_\text{ref} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation}

@@

This is the criterion of our CE identification.  The constant terms were determined
through the implementation of an empirical method with manual inspection of the CE
identification results on selected days of the whole dataset.  We tested seven sets
of days with different characteristics relevant to the efficiency of the
identification threshold.  These sets were random groups of about 20 -- 30 days with
the following characteristics:
(a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance,
(c) absence of clouds (by implementing a clear sky identification algorithm as discussed in @Natsis2023),
(d) absence of clouds and absence of EC events,
(e) with at least $60\,\%$ of the day length without clouds and presence of EC events,
(h) randomly selected days, and
(i) manually selected days.
For the latter case and where needed, we also used images from the sky-camera to
further aid the decision of the manual inspection.

@@

The definition of the CE events with this method has a degree of subjectivity, since
the actual clear sky irradiance is not known and can only be approximated.  However,
this method was proven capable in detecting all major CE events.  Where some CE
events with very low OIR may be not detected, these are few with small
over-irradiance, and it is unlikely that will affect significantly our results.

@@

```{r example-day, echo=FALSE, fig.cap="Example of CE identification in Thessaloniki for 2019-07-11. The green line with blue symbols depicts the measured GHI in one minute steps. The red line shows the modelled threshold $E$ for the detection of CE events, which are denoted with red circles. The black curve represents the TSI at the top of the atmosphere, adjusted for the actual Sun-Earth distance and multiplied by the cosine of SZA.", fig.pos='H'}
knitr::include_graphics("../images/example-days-18.png")
```

```{r example-year, echo=FALSE, fig.cap="Example scatter plot of the measured GHI and the reference clear sky irradiance in Thessaloniki for the year 2005. The over-irradiance for CE and ECE events is color coded, while the remaining data points are shown in black.", fig.pos='H'}
knitr::include_graphics("../images/P-example-years-13.png")
```

@@

@@

One aspect of this study is to investigate the time evolution of the CE events by
analyzing the GHI measurements at Thessaloniki. Cloud enhancements can be influenced
by different factors, such as the geometry, size and optical thickness of clouds,
their height in the atmosphere and local weather regimes [@Mol2023; @Veerman2022;
@Gristey2022; @Tzoumanikas2016].  Some of these factors are related to changes in
climate; hence it would be reasonable to expect contributing to the frequency of
occurrence of CE events over Thessaloniki, as well as to the average OIR and excess
irradiation.  The long-term trends were calculated using a first-order autoregressive
model with the 'maximum likelihood' fitting method [@Gardner1980; @Jones1980], by
implementing the function 'arima' from the library 'stats' of the R programming
language [@RCT2023]. All trends are reported together with their $2\sigma$ error.

@@

We have to note that the excess irradiation related to the CE events cannot be
directly linked to the total energy balance of the atmosphere. The net solar
radiation of the region is not increased, but is rather redistributed through the CE
events.  This is also depicted by the ECE irradiance values, which can exceed the
equivalent clear sky irradiance by a significant amount.

@@

Next we investigated the distribution of the CE events within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the monthly box and
whisker plot of the CE number of occurrences normalized with the highest median
value, which occurs in June, depicting a clear seasonal cycle.  Although CE events
are present throughout the year, the most active months are May and June. During the
winter (December -- February), the number of CE cases is about $25\,\%$ of the
maximum, while in the intermediate months, the number of occurrences gradually ramps
between the maximum and minimum.  This seasonality is a combined effect of different
factors, among them the types of clouds, their frequency of occurrence, the
seasonally varying relative position of the sun, as well as the local landscape
characteristics that may influence the formation of the clouds.  Unfortunately, lack
of detailed data on cloud formation, type and location is not allowing further
analysis. The interannual variability of the monthly CE events is quite high as
manifested by the large monthly extremes, especially in the summer.

```{r relative-month-occurrences, echo=FALSE, fig.cap="Seasonal variability of the number of CE events in Thessaloniki for the period 1992 -- 2023 normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have also been normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\\,\\%$ and the upper $75\\,\\%$ percentiles, the thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values."}
knitr::include_graphics("../images/clim-CE-month-norm-MAX-median-N-MW-1.png")
```

@@

```{r ovir-distribution, echo=FALSE, fig.cap="Relative frequency distribution of CE events in Thessaloniki for the period 1992 -- 2023 as a function of OIR."}
knitr::include_graphics("../images/P-relative-distribution-diff-1.png")
```

@@

The duration of the CE events is variable and can last for several minutes or even
more than an hour. In order to study the characteristics of these consecutive events,
we grouped them into bins of increasing duration in steps of one minute.  We have
identified `r ST_G0[, .N]` groups of CE in the whole period of study, where the group
of the longest duration of
`r ST_G0[, max(GLB_diff.N)]` minutes occurred on
`r strftime(ST_G0[which.max(GLB_diff.N), Date], "%d\\ %B %Y")`.
Figure\nobreakspace{}\ref{fig:ceg-duration-distribution} shows the frequency
distribution of the CE events according to their duration.  We conclude that although
some groups of events last for more than an hour, about $80\,\%$ have duration of
less than 5 minutes.

```{r ceg-duration-distribution, echo=FALSE, fig.cap="Relative frequency distribution of CE groups of consequent CE events according to their duration for Thessaloniki in the period 1992 -- 2023."}
knitr::include_graphics("../images/groups-1.png")
```

The relation between the duration and the mean OIR of the groups has also been
studied (Figure\nobreakspace{}\ref{fig:group-2d}).  Evidently, evens of high excess
irradiation have small duration and vice versa. The vast majority of grouped events
are associated with small excess irradiation (e.g., $<5\,\text{kJ}/\text{m}^2$) and
small duration (e.g., $<5\,\text{min}$) while groups with strong excess irradiation
and long duration are very rare.  Similar results of this relation have been reported
by @Zhang2018 in a study using a far higher sampling rate ($100\,\text{Hz}$) than
ours.

```{r group-2d, echo=FALSE, fig.cap="Relation of excess irradiation of CE groups with their duration for Thessaloniki in the period 1992 -- 2023. The logarithmic color scale denotes the frequency of the respective groups of events."}
knitr::include_graphics("../images/P-groups-bin2d-1.png")
```

@@

An aspect of the CE events that is commonly reported and has some significance on the
solar energy production infrastructure are the extreme CE events (ECE), where solar
irradiance exceeds the expected irradiance on top of the atmosphere
(Equation\nobreakspace{}\ref{eq:ECE}).  Analogous to
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} we have computed the
distribution of the number of occurrences of ECE events by month in
Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}. The most active period
for ECE events is in spring and early summer (March -- June), followed by a
period in late autumn (September and October).  This is probably related to the
weather characteristics in these periods, with frequent alternations between clear
sky periods and clouds.

```{r relative-month-occurancies-ECE, echo=FALSE, fig.cap="Seasonal variability of the number of ECE events in Thessaloniki for the period 1993 -- 2023 normalized to the maximum occurring in March, in the form of a box and whisker plot. The box contains the data between the lower $25\\,\\%$ and the upper $75\\,\\%$ percentiles. The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values."}
knitr::include_graphics("../images/clim-ECE-month-norm-MAX-median-N-2.png")
```

@@

The distribution of the ECE events
(Figure\nobreakspace{}\ref{fig:P-extreme-distribution}) shows that in rare cases the
OIR can exceed the TSI even by more than $400\,\text{W}/\text{m}^2$, while in
$75\,\%$ of the cases the OIR is below $200\,\text{W}/\text{m}^2$. The OIR for the
most frequent ECEs ranges between $140$ and $180\,\text{W}/\text{m}^2$.  These
findings are in accordance with the results of @Vamvakas2020, the only difference
being that the OIR values reported for are higher than those for Thessaloniki.

```{r P-extreme-distribution, echo=FALSE, fig.cap="Distribution of ECE events in Thessaloniki for the period 1992 -- 2023."}
knitr::include_graphics("../images/P-extreme-distribution-1.png")
```

@@

By creating a clear sky approximation representing the long- and short-term
variability of the expected clear sky GHI, we were able to identify cases of CE
events in Thessaloniki for the period 1993 -- 2023.  After analyzing the CE cases, we
found an increase of
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the mean annual irradiation of the CE events increasing with a rate of
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The most active months of CE events are May and June.  We found that continuous CE
events can last up to $140$ minutes, while the duration of $80\,\%$ of them is bellow
$5$ minutes.

We have observed ECE cases exceeding the TSI by $400\,\text{W}/\text{m}^{2}$ with
$75\,\%$ of the cases under $200\,\text{W}/\text{m}^{2}$. The climatological
characteristics of the ECE events showed that the most active months are spread in
half of the year and particularly in the periods March -- June, September and
October.  The magnitude of the ECE events identified in Thessaloniki events does not
exceed the values reported for sites with more favourable conditions for the
phenomenon [e.g. @Cordero2023]. Some of the characteristics of CE and ECE events we
analyzed have strong similarities with the results of @Vamvakas2020 for the city of
Patras, south of Thessaloniki, albeit with differences in the magnitude of OIR.

An interpretation of the CE trends shows that the interaction of GHI with the clouds,
through this 30-year period, is a dynamic phenomenon that needs further
investigation. For such studies it would be essential to have more detailed
information on cloud characteristics, especially in order to investigate possible
associations of the observed trends with changes in climate.
